---
title: "Gapminder dataset Exploration"
author: "Vetle Birkeland Huglen"
date: "24 9 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gapminder)
library(tidyverse)
```

##Smell test the data

Firstly, we will look at the dataset as a whole, and try to figure out the format of the dataset. 

```{r}
typeof(gapminder)
class(gapminder)
storage.mode(gapminder)
```

The `typeof()` and `storage.mode()` functions regard the **gapminder** dataset as lists, which can essentially be thought of as simply a collection of elements. The `class()`
 function reveals that the dataset is also a **data.frame**, which imposes further restrictions on the list, namely that two different variables can't have the same name, all elements are vectors, and all elements have equal length.
 
```{r}
nrow(gapminder)
ncol(gapminder)
dim(gapminder)
length(gapminder)
```

Above are a few different ways to access the size of the dataframe; `ncols()` and `nrows()` returns the number of columns and rows, `dim()` returns the dimensions as a vector and `length()` also returns the length of the object (in this case the number of columns). One can imagine that one should be careful using `length()` on objects with more than one dimension, as it is not intuitive from the name of the function which dimension it returns. At the same time, it is very convenient to use it on 1d vectors, especially if it is not clear if the vector is transposed. 


```{r}
str(gapminder)
```

The variables `country` and `continent` are factor-variables with a certain amount of levels (one factor per country/continent). `year` and `pop` are integers and `lifeExp` and `gdpPercap` are numeric values.

##Explore individual variables

```{r}
country <- gapminder$country
unique(country)
table(country)
```
Firstly we take a look into the `country` variable. `unique()` lists all uniques entries of input data, which in this case just is the list of countries. `table()` lists the counts of each unique entry, a function which can be very usefull. In this case, `summary()` would actually do the same as `table()`, because its output depends on the format of the input data. 

Now lets look at a numerical value, like `pop`:

```{r}
summary(gapminder$pop)
```
Here we see that the summary function instead returns information about the distribution.
We can create a density plot to better visualize the information gained. 

```{r}
p <-ggplot(gapminder, aes(pop)) +
  geom_density(fill = "pink")
p
```
This plot was quite hard to read, because of the logarithmic scale of the data. Lets try to fix that:

```{r}
p2 <-ggplot(gapminder, aes(pop)) +
  geom_density(fill="yellow") +
  scale_x_log10()
p2
```
Now its easy to see that the distribution of populations matches the output of the summary function.

## AND NOW LETS MAKE EVEN MORE PLOTS!!!


